---
title: "ESE527 Tuvalu"
author: "Huanyue Liao"
date: "2023-03-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(forecast) 
#install.packages("fpp2")
library(fpp2)
library(lmtest)
```

# Data Preprocessing

```{r}
file_name1 = "monthly sea level.txt"
file_name2 = "monthly air temp.txt"
file_name3 = "monthly water temp.txt"


sl_monthly_df = read.csv(file_name1, header = TRUE, sep = "", skip = 6)
at_monthly_df = read.csv(file_name2, header = TRUE, sep = "", skip = 6)
wt_monthly_df = read.csv(file_name3, header = TRUE, sep = "", skip = 6)


sl_monthly_df
at_monthly_df
wt_monthly_df
```

## 1.Clear Dataset:

### - Find NaN data and omit them

```{r}
## find NaN data
find_na <- function(dataset, method) {
  nullRows = is.na(dataset['Mean'])&is.na(dataset['Maximum'])&is.na(dataset['Minimum'])
  omit_dataset = dataset[!nullRows,]
  if (method == "findNA") return (dataset[nullRows,])
  if (method == "omitNA") return (omit_dataset)
}

sl_missing = find_na(sl_monthly_df,"findNA")[c("Year", "Mth")]
at_missing = find_na(at_monthly_df,"findNA")[c("Year", "Mth")]
wt_missing = find_na(wt_monthly_df,"findNA")[c("Year", "Mth")]

## write missing dataframe into Excel
#library(openxlsx)
#names <- list('Sea Level' = sl_missing, 'Air Temp' = at_missing, 'Water Temp' = wt_missing)
#write.xlsx(names, file = 'missing data.xlsx')

## omit NaN data
sl_monthly_df = find_na(sl_monthly_df,"omitNA")
at_monthly_df = find_na(at_monthly_df,"omitNA")
wt_monthly_df = find_na(wt_monthly_df,"omitNA")

sl_monthly_df
at_monthly_df
wt_monthly_df
```

### - Consider about El-Niño

**Strong El-Niño: 1997-1998,**

**Weaker El-Niño: 2002-2003.**

El Niño and La Niña are **climate patterns** in the Pacific Ocean that can affect weather worldwide. El-Niño condition has a strong effect on marine life off the Pacific coast. So we want to create a new dataset to store the dataset

```{r}
elnino_years = c(1997,1998)
omit_elnino <- function(dataset, years) {
  return(dataset[!(dataset$Year %in% years),])
}
omit_elnino2 <- function(dataset, yr) {
  return(dataset[(dataset$Year > yr),])
}

sl_monthly_df_elnino = omit_elnino2(sl_monthly_df, 1998)
at_monthly_df_elnino =  omit_elnino2(at_monthly_df, 1998)
wt_monthly_df_elnino =  omit_elnino2(wt_monthly_df, 1998)

sl_monthly_df_elnino
at_monthly_df_elnino
wt_monthly_df_elnino
```

### - Transform 'Mean' values of datasets into time series(ts):

### - Use the Kalman Filter to fill the missing data

-   Ignore the missing values
-   Filling in the missing values: expand your date index to include the missing observations, and use `na.approx` from `zoo` to fill in the missing values via interpolation.

```{r}
## Filling the missing data in dataframe df
# df <- data.frame{Date:xxxx, Mean:xxxx}
# type: "raw": return zoo ts with NaN; "filled": return zoo ts with replacement.
#       "kalman": return Missing Value Imputation by Kalman Smoothing and State Space Models

fill_zoo <- function(df, type) {
  alldates <- seq.Date(
  min(df$Date), 
  max(df$Date),
  "month"
  )
  allvalues <- merge(
    x=data.frame(Date=alldates),
    y=df,
    all.x=TRUE
  )
  zoovalues <- zoo(allvalues$Mean,allvalues$Date)
  approxvalues <- na.approx(zoovalues)
  kalmanvalues <- na_kalman(zoovalues, model = "StructTS", smooth = TRUE, nit = -1, maxgap = Inf)
  if (type == "raw") return (zoovalues)
  if (type == "filled") return(approxvalues)
  if (type == "kalman") return (kalmanvalues)
}
```

```{r}
#install.packages('imputeTS')
library(xts)
library(zoo)
library(imputeTS)
## clear NaN dataset
sl_monthly_df$Date <- as.Date(as.yearmon(paste(sl_monthly_df$Year, sl_monthly_df$Mth, sep = "-")))
at_monthly_df$Date <- as.Date(as.yearmon(paste(at_monthly_df$Year, at_monthly_df$Mth, sep = "-")))
wt_monthly_df$Date <- as.Date(as.yearmon(paste(wt_monthly_df$Year, wt_monthly_df$Mth, sep = "-")))

## Omit Ei Nino dataset
sl_monthly_df_elnino$Date <- as.Date(as.yearmon(paste(sl_monthly_df_elnino$Year, sl_monthly_df_elnino$Mth, sep = "-")))
at_monthly_df_elnino$Date <- as.Date(as.yearmon(paste(at_monthly_df_elnino$Year, at_monthly_df_elnino$Mth, sep = "-")))
wt_monthly_df_elnino$Date <- as.Date(as.yearmon(paste(wt_monthly_df_elnino$Year, wt_monthly_df_elnino$Mth, sep = "-")))


# Create new dataframes only have two columns: Date, Mean
sl_mon_mean = sl_monthly_df[c("Date", "Mean")]
at_mon_mean = at_monthly_df[c("Date", "Mean")]
wt_mon_mean = wt_monthly_df[c("Date", "Mean")]
sl_mon_el_mean = sl_monthly_df_elnino[c("Date", "Mean")]
at_mon_el_mean = at_monthly_df_elnino[c("Date", "Mean")]
wt_mon_el_mean = wt_monthly_df_elnino[c("Date", "Mean")]

## Raw time series data
raw_sl_mon = fill_zoo(sl_mon_mean, "raw")
raw_at_mon = fill_zoo(at_mon_mean, "raw")
raw_wt_mon = fill_zoo(wt_mon_mean, "raw")
raw_sl_mon_el = fill_zoo(sl_mon_el_mean, "raw")
raw_at_mon_el = fill_zoo(at_mon_el_mean, "raw")
raw_wt_mon_el = fill_zoo(wt_mon_el_mean, "raw")

## Filling time series data
approx_sl_mon = fill_zoo(sl_mon_mean, "filled")
approx_at_mon = fill_zoo(at_mon_mean, "filled")
approx_wt_mon = fill_zoo(wt_mon_mean, "filled")
approx_sl_mon_el = fill_zoo(sl_mon_el_mean, "filled")
approx_at_mon_el = fill_zoo(at_mon_el_mean, "filled")
approx_wt_mon_el = fill_zoo(wt_mon_el_mean, "filled")

## Filling time series data with Kalman filter
approx_sl_mon_el2 = fill_zoo(sl_mon_el_mean, "kalman")
approx_at_mon_el2 = fill_zoo(at_mon_el_mean, "kalman")
approx_wt_mon_el2 = fill_zoo(wt_mon_el_mean, "kalman")

approx_sl_mon_ts = ts(approx_sl_mon, start=c(1993,3), frequency = 12)
approx_at_mon_ts = ts(approx_at_mon, start=c(1993,3), frequency = 12)
approx_wt_mon_ts = ts(approx_wt_mon, start=c(1993,3), frequency = 12)
approx_sl_mon_el_ts = ts(approx_sl_mon_el, start=c(1999,1), frequency = 12)
approx_at_mon_el_ts = ts(approx_at_mon_el, start=c(1999,1), frequency = 12)
approx_wt_mon_el_ts = ts(approx_wt_mon_el, start=c(1999,1), frequency = 12)
approx_sl_mon_el_ts2 = ts(approx_sl_mon_el2, start=c(1999,1), frequency = 12)
approx_at_mon_el_ts2 = ts(approx_at_mon_el2, start=c(1999,1), frequency = 12)
approx_wt_mon_el_ts2 = ts(approx_wt_mon_el2, start=c(1999,1), frequency = 12)

#plot(na.interp(raw_sl_mon_el, lambda = 'auto'))
plot(approx_sl_mon_el_ts, main="Simple filling: sea level")
plot(approx_at_mon_el_ts, main="Simple filling: air temp ")
plot(approx_wt_mon_el_ts, main="Simple filling: water temp")
plot(approx_sl_mon_el_ts2, main="Kalman filling: sea level")
plot(approx_at_mon_el_ts2, main="Kalman filling: air temp ")
plot(approx_wt_mon_el_ts2, main="Kalman filling: water temp")
```

#### Raw Plots

```{r}
plot_ts <- function(tsdat, title, ylab, xlab) {
  autoplot(tsdat) +
    ggtitle(title) +
    ylab(ylab) +
    xlab(xlab)
    
}
plot_ts(raw_sl_mon, "Raw Monthly Sea Level Data", "Meter", "Year")
plot_ts(raw_at_mon, "Raw Monthly Air Temperature Data", "Celsius (°C)", "Year")
plot_ts(raw_wt_mon, "Raw Monthly Water Temperature Data", "Celsius (°C)", "Year")
plot_ts(raw_sl_mon_el, "Raw Monthly Sea Level Data Except Strong El-Niño", "Meter", "Year")
plot_ts(raw_at_mon_el, "Raw Monthly Air Temperature Data Except Strong El-Niño", "Celsius (°C)", "Year")
plot_ts(raw_wt_mon_el, "Raw Monthly Water Temperature Data Except Strong El-Niño", "Celsius (°C)", "Year")
```

#### Filled data plots

```{r}
plot_ts(approx_sl_mon_el2, "Preprocssing Monthly Sea Level Data Except Strong El-Niño", "Meter", "Year")
plot_ts(approx_at_mon_el2, "Preprocssing Monthly Air Temperature Data Except Strong El-Niño", "Celsius (°C)", "Year")
plot_ts(approx_wt_mon_el2, "Preprocssing Monthly Water Temperature Data Except Strong El-Niño", "Celsius (°C)", "Year")
```

## 2. Check if a time series is stationary: confirm the stationarity of the dataset

-   We use Dickey-Fuller test using `adf.test` function of `tseries` package. Augmented Dickey-Fuller (ADF) t-statistic test to find if the series has a unit root (a series with a trend line will have a unit root and result in a large p-value).

```{r}
options(warn=-1)
library(tseries)

adf.test(approx_sl_mon_el_ts2)

adf.test(approx_at_mon_el_ts2)

adf.test(approx_wt_mon_el_ts2)
```

### Check ACF

-   Autocorrelation Function (ACF) Identify if correlation at different time lags goes to 0. Just as correlation measures the extent of a linear relationship between two variables, autocorrelation measures the linear relationship between lagged values of a time series.

```{r}
library(cowplot)
#library(patchwork)
iris1 <- autoplot(approx_sl_mon_ts)+
  ggtitle("Monthly Sea Level Data") +
  ylab("meter") +
  xlab("Year")

## Seasonality Plot
#ggseasonplot(approx_sl_mon_ts, year.labels=TRUE, year.labels.left=TRUE) +
#  ylab("meter") +
#  ggtitle("Seasonal plot: Monthly Sea Level")


ggAcf(approx_sl_mon_el_ts2)+
  ggtitle("Autocorrelation for monthly sea level (m) except Strong El-Niño")

ggAcf(approx_at_mon_el_ts2)+
  ggtitle("Autocorrelation for monthly air temperature(°C) except Strong El-Niño")

ggAcf(approx_wt_mon_el_ts2)+
  ggtitle("Autocorrelation for monthly water temperature(°C) except Strong El-Niño")
```

Monthly sea level is not stationary, since most lags exceed the confidence interval of the ACF (blue dashed line). \## 3. Forecsating Model: Training&Test Datasets The study devide data into two groups, they are training data and testing data. The training data is a set of data that will be used to perform analysis and determine the model. Training data is data about product demand between 1st period and (n-12)th period where n is total period of data. The testing data is a set of data that will be used to test the accuracy of the forecast results. Testing data is about product demand between (n-11)th period and nth period where n is the total data.

```{r}
## training dataset: 4/5=232
## test dataset: the last 1/5 =290/5=58
split_train_test <- function(ts, type) {
  monthly = 1/12
  len = length(ts)
  n.train = as.integer(len*4/5)
  if (type=="train") return(window(ts, start=1999, end=1999+monthly*n.train-0.01))
  else if (type=="test") return(window(ts, start=1999+monthly*n.train))
}

## Train: 1999/01 to 2018/04; Test:2018/05 to 2023/02
sl_train = split_train_test(approx_sl_mon_el_ts2,"train")
sl_test = split_train_test(approx_sl_mon_el_ts2,"test")
at_train = split_train_test(approx_at_mon_el_ts2,"train")
at_test = split_train_test(approx_at_mon_el_ts2,"test")
wt_train = split_train_test(approx_wt_mon_el_ts2,"train")
wt_test = split_train_test(approx_wt_mon_el_ts2,"test")
```

## 3. STL model

-   t.window controls wiggliness of trend component.

-   s.window controls variation on seasonal component. These control how rapidly the trend-cycle and seasonal components can change. Smaller values allow for more rapid changes.

```{r stlagain, echo=TRUE, warning=FALSE, fig.width=8, fig.height=4}
library(Rsafd)
library(forecast)
library(ggplot2)
sl_el_fit <- stl(sl_train, s.window=12, robust=TRUE)

autoplot(sl_el_fit) +
  ggtitle("STL decomposition of monthly sea level (m) except strong El Niño.")

at_el_fit <- stl(at_train, s.window=12, robust=TRUE)
autoplot(at_el_fit) +
  ggtitle("STL decomposition of monthly air temperature (°C) except strong El Niño.")

wt_el_fit <- stl(wt_train, s.window=12, robust=TRUE)
autoplot(wt_el_fit) +
  ggtitle("STL decomposition of monthly water temperature (°C) except strong El Niño.")
```

### Find the outliers by using STL decomposition:

```{r}
plot_resid <- function(res) {
  mu = mean(res)
  sd = sd(res)
  plot(res) + 
  abline(h = c(mu-3*sd, mu+3*sd), col = c("blue","blue"), lty=c(2,2))
}
```

```{r}
## calculate the reminders in each STL decomposition
sl_el_resid <- remainder(sl_el_fit)
at_el_resid <- remainder(at_el_fit)
wt_el_resid <- remainder(wt_el_fit)

plot_resid(sl_el_resid)
plot_resid(at_el_resid)
plot_resid(wt_el_resid)
```

```{r}
tsoutliers(raw_sl_mon_el)
tsoutliers(raw_at_mon_el)
tsoutliers(raw_wt_mon_el)

tsoutliers(na.approx(raw_sl_mon_el))
tsoutliers(na.approx(raw_at_mon_el))
tsoutliers(na.approx(raw_wt_mon_el))
```

### Analyze stationary for residual first:

Analysis of the remainder part in order to find the most appropriate model for it. By adf.test, we know that the remainder parts are stationary time series. As the first step, we produce the auto-correlation plot using the R command acf.

If stationary: use AR/ARMA (Auto-regressive model) (auto-regressive moving average model)

**Why white noise residuals are desired:**

It is a general principle of time series models (including ARMA) that you would like to capture all systematic dynamics in the data. This means, every variation that can be explained by input variables, auto-regression, moving-average etc. should be explained.

If after your model building, the residuals are purely random, and are not correlated with each other i.e. show no systematic pattern, then you are generally satisfied, because you reached a point where you cannot go further. White noise is completely unpredictable.

**And a bit more pragmatic answer:**

The residuals of an ARMA process are by definition white noise. If you fit your data to an ARMA model, and you see that residuals are white noise, then this indicates you that the model fits the data well.

**Ljung-Box test for ARMA model residual diagnostics**:

For both Q and Q∗, the results are not significant (i.e. the p-values are relatively large). Thus, we can conclude that the residuals are not distinguishable from a white noise series. - **The Ljung-Box test uses the following hypotheses:**

-   **H0: The residuals are independently distributed.**

-   **HA: The residuals are not independently distributed; they exhibit serial correlation.**

    **Rejection Region**

    The test statistic *Q\** follows a chi-square distribution with *h* degrees of freedom; that is, Q\*\~ X^2^(h).

    We reject the null hypothesis and say that the residuals of the model are not independently distributed if Q\* \>X^2^~1-α,h~

Ideally, we would like to be unable to reject the null hypothesis. That is, we would like to see a *p-value greater than 0.05* because this means the residuals for our time series model are independent, which is often an assumption we make when creating a model. And a small value of Q\*, which suggests that the auto correlations come from a WN series. The large values of Q∗ suggest that the autocorrelations do not come from a white noise series (WN). The p-value is the probability of getting a value as large as or larger than that observed under the null hypothesis that the true innovations are independent. Therefore a small p-value is an evidence against independence.

```{r}
## check stationary
adf.test(sl_el_resid)
adf.test(at_el_resid)
adf.test(wt_el_resid)

## check acf
acf(sl_el_resid)
acf(at_el_resid)
acf(wt_el_resid)
```

From ADF test, the residuals are stationary. • The plot shows significant correlations in the remainder part of data. It also seems that it does not vanish after a finite number of steps so that we do not attempt to fit a MA(q) model but try an ARMA(p,q) model instead. Except air temperature, we use the ARIMA model.

```{r}
## fit the residuals into ARMA model to check the new residual series is whether WN or not
arma.fit.sl.resid <- auto.arima(sl_el_resid, max.order = 12, max.d=0, ic="aic", trace = T)
## ARIMA(2,0,1) -> ARMA(2,1)
plot(residuals(arma.fit.sl.resid), main="Residuals for ARMA(2,1) of sea level")


arma.fit.at.resid <- auto.arima(at_el_resid, max.order = 12, ic="aic", trace = T)
## ARIMA(1,0,0) 
plot(residuals(arma.fit.at.resid), main="Residuals for ARMA(1,0) of air temperature")


arma.fit.wt.resid <- auto.arima(wt_el_resid, max.order = 12, max.d=0, ic="aic", trace = T)
## ARIMA(2,0,1) -> ARMA(2,1)
plot(residuals(arma.fit.wt.resid), main="Residuals for ARMA(2,1) of water temperature")


checkresiduals(arma.fit.sl.resid)
checkresiduals(arma.fit.at.resid)
checkresiduals(arma.fit.wt.resid)
```

From the Q\* values and large p-values(all \>0.05), we can conclude that Sea level/ Water Temperature/ Air Temperature STL residuals series are not distinguishable from a white noise series (WN), which means residuals are independently distributed. Therefore, we don't need some future decomposition for residuals.

#### STL Plots

```{r stlagain, echo=TRUE, warning=FALSE, fig.width=8, fig.height=4}
#install.packages("stlplus")
library(stlplus)
# stlplus accepts Na ts data
alldates <- seq.Date(
  min(sl_monthly_df_elnino$Date), 
  sl_monthly_df_elnino$Date[length(sl_train)-1], #232
  "month"
  )
stlplus_plot <- function(tsdat, n_s, title, xlab, ylab, res) {
  monthNames <- c("Ja", "F", "Mr", "Ap", "Ma", "Jn", "Jl", "Au", "S", "O", "N", "D")
  # the important parameter is n.p, the number of measurements in a full period of   seasonal behavior. Since we are looking for a yearly effect, n.p should equal 12.
  stlMonthly <- stlplus(tsdat, t=alldates, n.p = 12, s.window=n_s)
  #, sub.start=1, sub.labels = monthNames
  
  if (res == "s") plot_seasonal(stlMonthly)
  else if (res == "t") plot_trend(stlMonthly)
  else if (res == "a") plot(stlMonthly, main=title, xlab=xlab, ylab=ylab)
}


stlplus_plot(sl_train, n_s=13, title="STL decomposition of monthly sea level (m) except strong El Niño.", xlab="Year", ylab="monthly sea level (m)", "a")

stlplus_plot(at_train, n_s=13, title="STL decomposition of monthly air temperature (°C) except strong El Niño.", xlab="Year", ylab="monthly air temperature (°C)", "a")

stlplus_plot(wt_train, n_s=13, title="STL decomposition of water temperature (°C) except strong El Niño.", xlab="Year", ylab="monthly water temperature (°C)", "a")
```

### Forecasting with STL decomposition

```{r}
sl_stl_pred <- forecast(sl_el_fit,method="naive", h=24)
at_stl_pred <- forecast(at_el_fit,method="naive", h=24)
wt_stl_pred <- forecast(wt_el_fit,method="naive", h=24)

plot(sl_stl_pred, xlab="Year", ylab="Sea Level (m)", main="Sea Level: Comparison between data train and prediction")
lines(approx_sl_mon_el_ts2, col='red')
abline(v=2018.333, lty=2, lwd=2)


plot(at_stl_pred, xlab="Year", ylab="Air Temperature (°C)", main="Air Temperature:Comparison between data train and prediction")
lines(approx_at_mon_el_ts2, col='red')
abline(v=2018.333, lty=2, lwd=2)

plot(wt_stl_pred, xlab="Year", ylab="Water Temperature (°C)", main="Water Temperature: Comparison between data train and prediction")
lines(approx_wt_mon_el_ts2, col='red')
abline(v=2018.333, lty=2, lwd=2)

```

```{r}
library(MLmetrics)
library(Metrics)
## Calculate MAPE(Mean Absolute Percentage Error)
cat("Sea Level STL MAPE: ", MAPE(sl_stl_pred$mean,sl_test),"\n")
cat("Air Temp STL MAPE: ", MAPE(at_stl_pred$mean,at_test),"\n")
cat("Water Temp STL MAPE: ", MAPE(wt_stl_pred$mean,wt_test),"\n")

## Calculate RMSE(Root Mean Square Error)
cat("Sea Level STL RMSE: ", rmse(sl_stl_pred$mean,sl_test),"\n")
cat("Air Temp STL RMSE: ", rmse(at_stl_pred$mean,at_test),"\n")
cat("Water Temp STL RMSE: ", rmse(wt_stl_pred$mean,wt_test),"\n")

summary(sl_stl_pred)
summary(at_stl_pred)
summary(wt_stl_pred)
```

##5. Plot a trendline on original data

```{r}
plot(sl_train, main="Monthly Sea Level Data (m)")
abline(reg=lm(sl_train~time(sl_train)))

plot(at_train, main="Monthly Air Temperature (°C)")
abline(reg=lm(at_train~time(at_train)))

plot(wt_train, main="Monthly Water Temperature (°C)")
abline(reg=lm(wt_train~time(wt_train)))
```

### Create a Box Plot by cycle

```{r}
boxplot(sl_train~cycle(sl_train), xlab="Date", ylab = "Sea Level", main = "Monthly sea level boxplot from 1999/01-2018/04")

boxplot(at_train~cycle(at_train), xlab="Date", ylab = "Air Temperature", main = "Monthly air temp boxplot from 1999/01-2018/04")

boxplot(wt_train~cycle(wt_train), xlab="Date", ylab = "Water Temperature", main = "Monthly water temp boxplot from 1999/01-2018/04")
```

From the above plot, you can see that the sea level goes higher in January, February, March, and April as compared to the other months of the years. \## 6. Seasonality plots

```{r}
sl_el_season <- sl_el_fit$time.series[,1]
at_el_season <- at_el_fit$time.series[,1] 
wt_el_season <- wt_el_fit$time.series[,1]


autoplot(sl_el_season, main="Seasonality of monthly sea level", ylab="sea level (m)")
autoplot(at_el_season, main="Seasonality of monthly air temperature", ylab="air temp (°C)")
autoplot(wt_el_season, main="Seasonality of monthly water temperature", ylab="water temp (°C)")
```
## (DELETE)6. Linear Regression model for sea level
```{r}
plot(approx_sl_mon_el_ts2, type = 'n', ylab="Tuvalu Sea Level (m)")
lines(sl_train)
lines(sl_test, col = "red")
abline(v = 2018.333, lty = 2, lwd = 2)
h = length(sl_test)

# Mean (based on the overall mean value)
f1 <- meanf(sl_train, h = h)
lines(f1$mean, lwd = 2, col = "yellow")

# Naive (based on the last value)
f2 <- rwf(sl_train, h = h)
lines(f2$mean, lwd = 2, col = "green")

# Drift (based on 1st and last value)
f3 <- rwf(sl_train, drift = TRUE, h = h)
lines(f3$mean, lwd = 2, col = "orange")

# Seasonal naive forecast
f4 <- snaive(sl_train, h = h)
lines(f4$mean, lwd = 2, col = "blue")

legend("bottomleft", c("Actual", "Mean", "Naive", "Drift", "Seasonal Naive"), lty=c(1,1,1,1,1), lwd=c(1,2,2,2,2), col=c("red","yellow","green","orange","blue"))
```

### 6.1 Estimations on test datasets
```{r}
mapes<-list()
rmses<-list()
# display accuracy for mean method
MAPE(f1$mean, sl_test) 
RMSE(f1$mean, sl_test)
mapes <- append(mapes,MAPE(f1$mean, sl_test))
rmses <- append(rmses,RMSE(f1$mean, sl_test))
# display accuracy for naive method
MAPE(f2$mean, sl_test) 
RMSE(f2$mean, sl_test) 
mapes <- append(mapes,MAPE(f2$mean, sl_test))
rmses <- append(rmses,RMSE(f2$mean, sl_test))
# display accuracy for drift method
MAPE(f3$mean, sl_test) 
RMSE(f3$mean, sl_test)
mapes <- append(mapes,MAPE(f3$mean, sl_test))
rmses <- append(rmses,RMSE(f3$mean, sl_test))
# display accuracy for seasonal method (BEST)
MAPE(f4$mean, sl_test) 
RMSE(f4$mean, sl_test) 
mapes <- append(mapes,MAPE(f4$mean, sl_test))
rmses <- append(rmses,RMSE(f4$mean, sl_test))

rbind(mapes,rmses)
```
COMMENT :seasonal naive method is the one that minimizes RMSE on test set with the value of 0.06352969 Our intuition on the seasonal structure of the data has been confirmed by accuracy measures. seasonal naive method is then the best method to use for our prediction task.
```{r}
res <- residuals(f4)
plot(res)
acf(res, na.action = na.omit)

fit<-tslm(sl_train~trend)
f<-forecast(fit, h=h)
plot(f)
acf(residuals(f))

fit2<-tslm(sl_train~trend+season)
f2<-forecast(fit2, h=h)
plot(f2)
acf(residuals(f2))
pacf(residuals(f2))
```


## 7. ARIMAX model

### 7.1 Inferring causality in time series data: determine if there ia any causal relationship between two trends🧸

### 7.1.1 Estimate a time series Y and check the p-value of parameters for ARIMAX model

ARIMAX Equation: $$P_t=c+\beta X+\Phi_1P_{t-1}+\theta_1\epsilon_{t-1}+\epsilon_t $$ where $\epsilon$ is error term, $\Phi_1$ and $\beta_1$ are coefficient parameters, $X$ is the additional term.
```{r}
sl_el_trend <- sl_el_fit$time.series[,2]
at_el_trend <- at_el_fit$time.series[,2] 
wt_el_trend <- wt_el_fit$time.series[,2]

sl_el_deseasonal <- seasadj(sl_el_fit)
at_el_deseasonal <- seasadj(at_el_fit)
wt_el_deseasonal <- seasadj(wt_el_fit)
```

```{r}
library(lmtest)
set.seed(5)
#length(sl_el_trend)=232
## Let P[t-1]=1, c=1, beta=1, Phi_1=1, theta_1=1
err <- rnorm(n=232, mean = 0, sd = 1)

#Y_causal <- numeric(3)
#Y_causal <- 1 + 2*wt_el_trend + 3 * sl_el_trend[1:231] + 4*err[1:231] + err[2:232]
Y_causal <- numeric(232)
Y_causal[1] = 1  
for (i in 2:232) {
  Y_causal[i] <- 0.1 + 4*wt_el_deseasonal[i] + 0.5 * Y_causal[i-1] + 2*err[i-1] + err[i]
}

arimax_causal <- auto.arima(Y_causal, xreg=wt_el_deseasonal,  seasonal = F)
arimax_causal_inverse <- auto.arima(wt_el_deseasonal, xreg=Y_causal,  seasonal = F)

coeftest(arimax_causal) #ARIMA(0,0,2)
coeftest(arimax_causal_inverse) #ARIMA(2,0,1)
arimax_causal
arimax_causal_inverse
```

### 7.2.1 $T_w(t) \rightarrow T_s(t)$


$$sea\ level: W(t)=T_s(t)+S_s(t)+R_s(t)$$ $$water\ temp:T_w(t)=T_w(t)+S_w(t)+R_w(t)$$ Finding causal relationship between two sets of time series data $T_s(t),T_w(t)$.

Suppose that $T_w(t) \rightarrow T_s(t)$, as same as $T_s(t)$\~$T_w(t)$.

```{r}
autoplot(sl_el_trend, main="Trend of monthly sea level", ylab="sea level (m)")
autoplot(at_el_trend, main="Trend of monthly air temperature", ylab="air temp (°C)")
autoplot(wt_el_trend, main="Trend of monthly water temperature", ylab="water temp (°C)")
```
```{r}
## Generate the list of A specification of the non-seasonal part of the ARIMA model: the three components (p, d, q)
generate_orderlist <- function(maxp,maxd,maxq) {
  len <- (maxp+1)*(maxq+1)*(maxd+1)
  ll <- matrix(nrow = len, ncol = 3)
  return (expand.grid(p = 0:maxp, d = 0:maxd, q = 0:maxq))
}

## ## Calculate MAPE(Mean Absolute Percentage Error)
mape_cv <- function(err_ts) {
  return (mean(abs(err_ts), na.rm=T))
}

## Calculate RMSE(Root Mean Square Error)
rmse_cv <- function(err_ts) {
  return (sqrt(mean(err_ts^2, na.rm=T)))
}
```


```{r}
#library(TSA)
library(forecast)
#arimax(x, order = c(0, 0, 0), seasonal = list(order = c(0, 0, 0), period = NA),xreg = NULL, include.mean = TRUE, transform.pars = TRUE, fixed = NULL, init = NULL, method = c("CSS-ML", "ML", "CSS"), n.cond, optim.control = list(),kappa = 1e+06, io = NULL, xtransf, transfer = NULL)


arimax_w_s_fit <- auto.arima(sl_el_deseasonal, xreg=wt_el_deseasonal, seasonal = F)
arimax_s_w_fit <- auto.arima(wt_el_deseasonal, xreg=sl_el_deseasonal, seasonal = F)

## ARIMAX: ARIMA(2,0,1), ARIMA(1,1,1)
coeftest(arimax_w_s_fit)
coeftest(arimax_s_w_fit)

acf(sl_el_deseasonal)
acf(sl_el_deseasonal, type="par")
arimax_orders = generate_orderlist(2,1,2)
arimax_orders
```


```{r}
## Fit an ARIMAX model to each rolling origin subset with exogenous predictors
#far_xreg <- function(x, h, xreg, newxreg) {
#  forecast(auto.arima(x, xreg=xreg), xreg=newxreg)
#}
orders <- list()
mapes_w_s <- list()
rmses_w_s <- list()
for (i in 1:nrow(arimax_orders)) {
  far2_xreg <- function(x, h, xreg, newxreg) {
    forecast(Arima(x, order=as.numeric(arimax_orders[i,]), xreg=xreg), xreg=newxreg)
  }
  e <- tsCV(sl_el_deseasonal, far2_xreg, h=1, window = 58, xreg=wt_el_deseasonal)
  w <- paste('(',toString(arimax_orders[i,]),')')
  orders <- append(orders, w)
  mapes_w_s <- append(mapes_w_s, mape_cv(e))
  rmses_w_s <- append(rmses_w_s, rmse_cv(e))
}

plot(1:18, mapes_w_s, ylim=c(0.03,0.07), type='l', ylab="", xlab="index of ARIMAX model", main="Rolling CV of ARIMAX for T_s(t)~T_w(t)")
lines(1:18, rmses_w_s, col="red")
legend("topright", c("MAPE","RMSE"), col = c("black","red"), lty=c(1,1))

rbind(orders,mapes_w_s, rmses_w_s)

## Find monthly average seasonal values for the water temp of training data
wt_seasonal_est <- tapply(wt_el_season,cycle(wt_el_season),mean)
## Estimate seasonal data for test dataset: 2018 May - 2023 Feb
wt_seasonal_test <- c(wt_seasonal_est[5:12], rep(wt_seasonal_est,4), wt_seasonal_est[1:2])
## Compute the deseasonal test dataset for water temp
wt_test_deseasonal <- wt_test-wt_seasonal_test

arimax_ws_pred <- forecast(Arima(sl_el_deseasonal, order=c(1,0,0),xreg=wt_el_deseasonal), xreg=wt_test_deseasonal, h=58)
arimax_ws_pred

## Calculate MAPE(Mean Absolute Percentage Error)
cat("T_s(s) ~ T_w(t) ARIMAX MAPE: ", MAPE(arimax_ws_pred$mean,sl_test[1:58]),"\n")

## Calculate RMSE(Root Mean Square Error)
cat("T_s(s) ~ T_w(t) ARIMAX RMSE: ", rmse(arimax_ws_pred$mean,sl_test[1:58]),"\n")

plot(arimax_ws_pred,xlab="Time", ylab="Sea Level(m)", main="ARIMAX model of Sea Level Trend", ylim=c(2,2.3), col='blue')
lines(sl_test, col='red')
abline(v=2018.333, lty=2, lwd=2)
legend("topleft", c("ARIMAX predicted", "Actual"), lty=c(1,1), col=c("blue", "red"), bty="n")

## Using ARIMAX trend model to forecast the sea level
pred_train1 <- arimax_ws_pred$fitted+sl_el_season
plot(pred_train1, col="blue", ylim=c(1.8,2.25), ylab="Sea level (m)", main="Forecast sea level by trend ARIMAX model")
lines(sl_train, col="red")
legend("topleft", c("ARIMAX predicted", "Actual"), lty=c(1,1), col=c("blue", "red"), bty="n")

## Calculate MAPE(Mean Absolute Percentage Error)
cat("T_s(s) ~ T_w(t) ARIMAX MAPE for training dataset sea level: ", MAPE(pred_train1,sl_train),"\n")

## Calculate RMSE(Root Mean Square Error)
cat("T_s(s) ~ T_w(t) ARIMAX RMSE for training dataset sea level: ", rmse(pred_train1,sl_train),"\n")
```

### 7.2.2 $T_a \rightarrow T_s$
### 7.2.2 Cross-validation for ARIMAX: Rolling Time Series Cross Validation

$$air\ temp:T_a(t)=T_a(t)+S_a(t)+R_a(t)$$ Finding causal relationship between two sets of time series data $T_s(t),T_a(t)$.

Suppose that $T_a(t) \rightarrow T_s(t)$, as same as $T_s(t)$\~$T_a(t)$.

```{r}
arimax_a_s_fit <- auto.arima(sl_el_deseasonal, xreg=at_el_deseasonal, seasonal = F)
arimax_s_a_fit <- auto.arima(at_el_deseasonal, xreg=sl_el_deseasonal, seasonal = F)

## ARIMAX: ARIMA(2,0,1), ARIMA(2,1,1)
coeftest(arimax_a_s_fit)
coeftest(arimax_s_a_fit) 

#arimax_as_pred <- forecast(arimax_a_s_fit, xreg=at_test, h=58)
#arimax_as_pred

## CV: Fit an ARIMAX model to each rolling origin subset with exogenous predictors
mapes_a_s <- list()
rmses_a_s <- list()
for (i in 1:nrow(arimax_orders)) {
  far2_xreg <- function(x, h, xreg, newxreg) {
    forecast(Arima(x, order=as.numeric(arimax_orders[i,]), xreg=xreg), xreg=newxreg)
  }
  e <- tsCV(sl_el_deseasonal, far2_xreg, h=1, window = 58, xreg=at_el_deseasonal)
  mapes_a_s <- append(mapes_a_s, mape_cv(e))
  rmses_a_s <- append(rmses_a_s, rmse_cv(e))
}

plot(1:18, mapes_a_s, ylim=c(0.03, 0.066), type='l', ylab="", xlab="index of ARIMAX model", main="Rolling CV of ARIMAX for T_a(t)~T_w(t)")
lines(1:18, rmses_a_s, col="red")
legend("topright", c("MAPE","RMSE"), col = c("black","red"), lty=c(1,1))

rbind(orders,mapes_a_s, rmses_a_s)

## Find monthly average seasonal values for the air temp of training data
at_seasonal_est <- tapply(at_el_season,cycle(at_el_season),mean)
## Estimate seasonal data for test dataset: 2018 May - 2023 Feb
at_seasonal_test <- c(wt_seasonal_est[5:12], rep(wt_seasonal_est,4), at_seasonal_est[1:2])
## Compute the deseasonal test dataset for water temp
at_test_deseasonal <- at_test-at_seasonal_test

arimax_as_pred <- forecast(Arima(sl_el_deseasonal, order=c(2,0,1), xreg=at_el_deseasonal), xreg=at_test_deseasonal, h=58)
arimax_as_pred
```

```{r}
## Calculate MAPE(Mean Absolute Percentage Error)
cat("T_s(s) ~ T_a(t) ARIMAX MAPE: ", MAPE(arimax_as_pred$mean,sl_test[1:58]),"\n")

## Calculate RMSE(Root Mean Square Error)
cat("T_s(s) ~ T_a(t) ARIMAX RMSE: ", rmse(arimax_as_pred$mean,sl_test[1:58]),"\n")

plot(arimax_as_pred,xlab="Time", ylab="Sea Level(m)", main="ARIMAX model of Sea Level Trend", ylim=c(2,2.3), col='blue')
lines(sl_test, col='red')
abline(v=2018.333, lty=2, lwd=2)
legend("topleft", c("ARIMAX predicted", "Actual"), lty=c(1,1), col=c("blue", "red"), bty="n")

## Using ARIMAX trend model to forecast the sea level
pred_train1 <- arimax_as_pred$fitted+sl_el_season
plot(pred_train1, col="blue", ylim=c(1.8,2.25), ylab="Sea level (m)", main="Forecast sea level by trend ARIMAX model")
lines(sl_train, col="red")
legend("topleft", c("ARIMAX predicted", "Actual"), lty=c(1,1), col=c("blue", "red"), bty="n")

## Calculate MAPE(Mean Absolute Percentage Error)
cat("T_s(s) ~ T_a(t) ARIMAX MAPE for training dataset sea level: ", MAPE(pred_train1,sl_train),"\n")

## Calculate RMSE(Root Mean Square Error)
cat("T_s(s) ~ T_a(t) ARIMAX RMSE for training dataset sea level: ", rmse(pred_train1,sl_train),"\n")
```

### 7.2.3 $T_a + T_w \rightarrow T_s$

```{r}
awt_el_deseasonal <- cbind(at_el_deseasonal,wt_el_deseasonal)
arimax_aw_s_fit <- auto.arima(sl_el_deseasonal, xreg=awt_el_deseasonal, seasonal = F)
## ARIMA(2,0,1)
coeftest(arimax_aw_s_fit)

#arimax_aws_pred <- forecast(arimax_aw_s_fit, xreg=cbind(at_test,wt_test), h=58)
#arimax_aws_pred

## CV: Fit an ARIMAX model to each rolling origin subset with exogenous predictors
mapes_aw_s <- list()
rmses_aw_s <- list()
for (i in 1:nrow(arimax_orders)) {
  far2_xreg <- function(x, h, xreg, newxreg) {
    forecast(Arima(x, order=as.numeric(arimax_orders[i,]), xreg=xreg), xreg=newxreg)
  }
  e <- tsCV(sl_el_deseasonal, far2_xreg, h=1, window = 58, xreg=awt_el_deseasonal)
  mapes_aw_s <- append(mapes_aw_s, mape_cv(e))
  rmses_aw_s <- append(rmses_aw_s, rmse_cv(e))
}

plot(1:18, mapes_aw_s, ylim=c(0.03, 0.066), type='l', ylab="", xlab="index of ARIMAX model", main="Rolling CV of ARIMAX for T_w(t)+T_a(t)~T_w(t)")
lines(1:18, rmses_aw_s, col="red")
legend("topright", c("MAPE","RMSE"), col = c("black","red"), lty=c(1,1))

rbind(orders,mapes_aw_s, rmses_aw_s)

arimax_aws_pred <- forecast(Arima(sl_el_deseasonal, order=c(1,0,0), xreg=awt_el_deseasonal), xreg=cbind(at_test_deseasonal,wt_test_deseasonal), h=58)
arimax_aws_pred
```

```{r}
## Calculate MAPE(Mean Absolute Percentage Error)
cat("T_s(s) ~ T_a(t)+T_w(t) ARIMAX MAPE: ", MAPE(arimax_aws_pred$mean,sl_test[1:58]),"\n")

## Calculate RMSE(Root Mean Square Error)
cat("T_s(s) ~ T_a(t)+T_w(t) ARIMAX RMSE: ", rmse(arimax_aws_pred$mean,sl_test[1:58]),"\n")

plot(arimax_aws_pred,xlab="Time", ylab="Sea Level (m)", main="ARIMAX model of Sea Level Trend", ylim=c(2,2.3), col='blue')
lines(sl_test, col='red')
abline(v=2018.333, lty=2, lwd=2)
legend("topleft", c("ARIMAX predicted", "Actual"), lty=c(1,1), col=c("blue", "red"), bty="n")

## Using ARIMAX trend model to forecast the sea level
pred_train3 <- arimax_aws_pred$fitted+sl_el_season
plot(pred_train3, col="blue", ylim=c(1.8,2.25), ylab="Sea level", main="Forecast sea level by trend ARIMAX model")
lines(sl_train, col="red")
legend("topleft", c("ARIMAX predicted", "Actual"), lty=c(1,1), col=c("blue", "red"), bty="n")

## Calculate MAPE(Mean Absolute Percentage Error)
cat("T_s(s) ~ T_a(t)+T_w(t) ARIMAX MAPE for training dataset sea level: ", MAPE(pred_train3,sl_train),"\n")

## Calculate RMSE(Root Mean Square Error)
cat("T_s(s) ~ T_a(t)+T_w(t) ARIMAX RMSE for training dataset sea level: ", rmse(pred_train3,sl_train),"\n")
```

## 8. Linear Regression on sea level trend
```{r}
plot(sl_el_trend)
abline(reg=lm(sl_el_trend~time(sl_el_trend)), col="red")
lmfit <- lm(sl_el_trend~time(sl_el_trend))
lm.slope <- coef(lmfit)[2]
cat("Sea level yearly increasing ", lm.slope*12," meter.\n")
```
