---
title: "ESE527 Tuvalu"
author: "Huanyue Liao"
date: "2023-03-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(forecast) 
#install.packages("fpp2")
library(fpp2)
```

# Data Preprocessing

```{r}
file_name1 = "monthly sea level.txt"
file_name2 = "monthly air temp.txt"
file_name3 = "monthly water temp.txt"
file_name4 = "dailySL1993-2018.csv"

sl_monthly_df = read.csv(file_name1, header = TRUE, sep = "", skip = 6)
sl_monthly_df
at_monthly_df = read.csv(file_name2, header = TRUE, sep = "", skip = 6)
wt_monthly_df = read.csv(file_name3, header = TRUE, sep = "", skip = 6)
sl_daily_df = read.csv(file_name4, header=FALSE, sep = ",")
```

## 1.Clear Dataset:

### - Find NaN data and omit them

```{r}
## find NaN data
find_na <- function(dataset, method) {
  nullRows = is.na(dataset['Mean'])&is.na(dataset['Maximum'])&is.na(dataset['Minimum'])
  omit_dataset = dataset[!nullRows,]
  if (method == "findNA") return (dataset[nullRows,])
  if (method == "omitNA") return (omit_dataset)
}

find_na(sl_monthly_df,"findNA")
find_na(at_monthly_df,"findNA")
find_na(wt_monthly_df,"findNA")

## omit NaN data
sl_monthly_df = find_na(sl_monthly_df,"omitNA")
at_monthly_df = find_na(at_monthly_df,"omitNA")
wt_monthly_df = find_na(wt_monthly_df,"omitNA")
```

### - Consider about  El-Niño

**Strong El-Niño: 1997-1998,**

**Weaker El-Niño: 2002-2003.**

El Niño and La Niña are **climate patterns** in the Pacific Ocean that can affect weather worldwide.
El-Niño condition has a strong effect on marine life off the Pacific coast. So we want to create a new dataset to store the dataset

```{r}
elnino_years = c(1997,1998)
omit_elnino <- function(dataset, years) {
  return(dataset[!(dataset$Year %in% years),])
}
sl_monthly_df_elnino = omit_elnino(sl_monthly_df, elnino_years)
at_monthly_df_elnino =  omit_elnino(at_monthly_df, elnino_years)
wt_monthly_df_elnino =  omit_elnino(wt_monthly_df, elnino_years)
```

### - Transform 'Mean' values of datasets into time series(ts):
- Ignore the missing values
- Filling in the missing values: expand your date index to include the missing observations, and use `na.approx` from `zoo` to fill in the missing values via interpolation.
```{r}
## Filling the missing data in dataframe df
# df <- data.frame{Date:xxxx, Mean:xxxx}
fill_zoo <- function(df) {
  alldates <- seq.Date(
  min(df$Date), 
  max(df$Date),
  "month"
  )
  allvalues <- merge(
    x=data.frame(Date=alldates),
    y=df,
    all.x=TRUE
  )
  zoovalues <- zoo(allvalues$Mean,allvalues$Date)
  approxvalues <- na.approx(zoovalues)
  return(approxvalues)
}
```

```{r}
library(xts)
library(zoo)
## clear NaN dataset
sl_monthly_df$Date <- as.Date(as.yearmon(paste(sl_monthly_df$Year, sl_monthly_df$Mth, sep = "-")))
at_monthly_df$Date <- as.Date(as.yearmon(paste(at_monthly_df$Year, at_monthly_df$Mth, sep = "-")))
wt_monthly_df$Date <- as.Date(as.yearmon(paste(wt_monthly_df$Year, wt_monthly_df$Mth, sep = "-")))

## Omit Ei Nino dataset
sl_monthly_df_elnino$Date <- as.Date(as.yearmon(paste(sl_monthly_df_elnino$Year, sl_monthly_df_elnino$Mth, sep = "-")))
at_monthly_df_elnino$Date <- as.Date(as.yearmon(paste(at_monthly_df_elnino$Year, at_monthly_df_elnino$Mth, sep = "-")))
wt_monthly_df_elnino$Date <- as.Date(as.yearmon(paste(wt_monthly_df_elnino$Year, wt_monthly_df_elnino$Mth, sep = "-")))


# Create new dataframes only have two columns: Date, Mean
sl_mon_mean = sl_monthly_df[c("Date", "Mean")]
at_mon_mean = at_monthly_df[c("Date", "Mean")]
wt_mon_mean = wt_monthly_df[c("Date", "Mean")]
sl_mon_el_mean = sl_monthly_df_elnino[c("Date", "Mean")]
at_mon_el_mean = at_monthly_df_elnino[c("Date", "Mean")]
wt_mon_el_mean = wt_monthly_df_elnino[c("Date", "Mean")]

approx_sl_mon = fill_zoo(sl_mon_mean)
approx_at_mon = fill_zoo(at_mon_mean)
approx_wt_mon = fill_zoo(wt_mon_mean)
approx_sl_mon_el = fill_zoo(sl_mon_el_mean)
approx_at_mon_el = fill_zoo(at_mon_el_mean)
approx_wt_mon_el = fill_zoo(wt_mon_el_mean)


approx_sl_mon_ts = ts(approx_sl_mon, start=c(1993,3), frequency = 12)
approx_at_mon_ts = ts(approx_at_mon, start=c(1993,3), frequency = 12)
approx_wt_mon_ts = ts(approx_wt_mon, start=c(1993,3), frequency = 12)
approx_sl_mon_el_ts = ts(approx_sl_mon_el, start=c(1993,3), frequency = 12)
approx_at_mon_el_ts = ts(approx_at_mon_el, start=c(1993,3), frequency = 12)
approx_wt_mon_el_ts = ts(approx_wt_mon_el, start=c(1993,3), frequency = 12)
```

## STL model
- t.window controls wiggliness of trend component.

- s.window controls variation on seasonal component.
```{r stlagain, echo=TRUE, warning=FALSE, fig.width=8, fig.height=4}
sl_fit <- stl(approx_sl_mon_ts, s.window=5, robust=TRUE)
autoplot(sl_fit) +
  ggtitle("STL decomposition of monthly sea level (m)")
```


```{r mstl, fig.width=8, fig.height=4}
approx_sl_mon_ts %>% mstl() %>% autoplot()
```
