---
title: "ESE527 Tuvalu"
author: "Huanyue Liao"
date: "2023-03-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(forecast) 
#install.packages("fpp2")
library(fpp2)
```

# Data Preprocessing

```{r}
file_name1 = "monthly sea level.txt"
file_name2 = "monthly air temp.txt"
file_name3 = "monthly water temp.txt"
file_name4 = "dailySL1993-2018.csv"
file_name5 = "dailySL1977-1999.csv"

sl_monthly_df = read.csv(file_name1, header = TRUE, sep = "", skip = 6)
sl_monthly_df
at_monthly_df = read.csv(file_name2, header = TRUE, sep = "", skip = 6)
wt_monthly_df = read.csv(file_name3, header = TRUE, sep = "", skip = 6)
sl_daily_df_93_18 = read.csv(file_name4, header=FALSE, sep = ",")
sl_daily_df_77_99 = read.csv(file_name5, header=FALSE, sep = ",")
sl_daily_df_93_18
sl_daily_df_77_99
```

## 1.Clear Dataset:

### - Find NaN data and omit them

```{r}
## find NaN data
find_na <- function(dataset, method) {
  nullRows = is.na(dataset['Mean'])&is.na(dataset['Maximum'])&is.na(dataset['Minimum'])
  omit_dataset = dataset[!nullRows,]
  if (method == "findNA") return (dataset[nullRows,])
  if (method == "omitNA") return (omit_dataset)
}

sl_missing = find_na(sl_monthly_df,"findNA")[c("Year", "Mth")]
at_missing = find_na(at_monthly_df,"findNA")[c("Year", "Mth")]
wt_missing = find_na(wt_monthly_df,"findNA")[c("Year", "Mth")]

## write missing dataframe into Excel
#library(openxlsx)
#names <- list('Sea Level' = sl_missing, 'Air Temp' = at_missing, 'Water Temp' = wt_missing)
#write.xlsx(names, file = 'missing data.xlsx')

## omit NaN data
sl_monthly_df = find_na(sl_monthly_df,"omitNA")
at_monthly_df = find_na(at_monthly_df,"omitNA")
wt_monthly_df = find_na(wt_monthly_df,"omitNA")

sl_monthly_df
at_monthly_df
wt_monthly_df
```

### - Consider about El-Niño

**Strong El-Niño: 1997-1998,**

**Weaker El-Niño: 2002-2003.**

El Niño and La Niña are **climate patterns** in the Pacific Ocean that can affect weather worldwide. El-Niño condition has a strong effect on marine life off the Pacific coast. So we want to create a new dataset to store the dataset

```{r}
elnino_years = c(1997,1998)
omit_elnino <- function(dataset, years) {
  return(dataset[!(dataset$Year %in% years),])
}
sl_monthly_df_elnino = omit_elnino(sl_monthly_df, elnino_years)
at_monthly_df_elnino =  omit_elnino(at_monthly_df, elnino_years)
wt_monthly_df_elnino =  omit_elnino(wt_monthly_df, elnino_years)

sl_monthly_df_elnino
at_monthly_df_elnino
wt_monthly_df_elnino
```

### - Transform 'Mean' values of datasets into time series(ts):

-   Ignore the missing values
-   Filling in the missing values: expand your date index to include the missing observations, and use `na.approx` from `zoo` to fill in the missing values via interpolation.

```{r}
## Filling the missing data in dataframe df
# df <- data.frame{Date:xxxx, Mean:xxxx}
# type: "raw": return zoo ts with NaN; "filled": return zoo ts with replacement.
fill_zoo <- function(df, type) {
  alldates <- seq.Date(
  min(df$Date), 
  max(df$Date),
  "month"
  )
  allvalues <- merge(
    x=data.frame(Date=alldates),
    y=df,
    all.x=TRUE
  )
  zoovalues <- zoo(allvalues$Mean,allvalues$Date)
  approxvalues <- na.approx(zoovalues)
  if (type == "raw") return (zoovalues)
  if (type == "filled") return(approxvalues)
}
```

```{r}
library(xts)
library(zoo)
## clear NaN dataset
sl_monthly_df$Date <- as.Date(as.yearmon(paste(sl_monthly_df$Year, sl_monthly_df$Mth, sep = "-")))
at_monthly_df$Date <- as.Date(as.yearmon(paste(at_monthly_df$Year, at_monthly_df$Mth, sep = "-")))
wt_monthly_df$Date <- as.Date(as.yearmon(paste(wt_monthly_df$Year, wt_monthly_df$Mth, sep = "-")))

## Omit Ei Nino dataset
sl_monthly_df_elnino$Date <- as.Date(as.yearmon(paste(sl_monthly_df_elnino$Year, sl_monthly_df_elnino$Mth, sep = "-")))
at_monthly_df_elnino$Date <- as.Date(as.yearmon(paste(at_monthly_df_elnino$Year, at_monthly_df_elnino$Mth, sep = "-")))
wt_monthly_df_elnino$Date <- as.Date(as.yearmon(paste(wt_monthly_df_elnino$Year, wt_monthly_df_elnino$Mth, sep = "-")))


# Create new dataframes only have two columns: Date, Mean
sl_mon_mean = sl_monthly_df[c("Date", "Mean")]
at_mon_mean = at_monthly_df[c("Date", "Mean")]
wt_mon_mean = wt_monthly_df[c("Date", "Mean")]
sl_mon_el_mean = sl_monthly_df_elnino[c("Date", "Mean")]
at_mon_el_mean = at_monthly_df_elnino[c("Date", "Mean")]
wt_mon_el_mean = wt_monthly_df_elnino[c("Date", "Mean")]

## Raw time series data
raw_sl_mon = fill_zoo(sl_mon_mean, "raw")
raw_at_mon = fill_zoo(at_mon_mean, "raw")
raw_wt_mon = fill_zoo(wt_mon_mean, "raw")
raw_sl_mon_el = fill_zoo(sl_mon_el_mean, "raw")
raw_at_mon_el = fill_zoo(at_mon_el_mean, "raw")
raw_wt_mon_el = fill_zoo(wt_mon_el_mean, "raw")

## Filling time series data
approx_sl_mon = fill_zoo(sl_mon_mean, "filled")
approx_at_mon = fill_zoo(at_mon_mean, "filled")
approx_wt_mon = fill_zoo(wt_mon_mean, "filled")
approx_sl_mon_el = fill_zoo(sl_mon_el_mean, "filled")
approx_at_mon_el = fill_zoo(at_mon_el_mean, "filled")
approx_wt_mon_el = fill_zoo(wt_mon_el_mean, "filled")


approx_sl_mon_ts = ts(approx_sl_mon, start=c(1993,3), frequency = 12)
approx_at_mon_ts = ts(approx_at_mon, start=c(1993,3), frequency = 12)
approx_wt_mon_ts = ts(approx_wt_mon, start=c(1993,3), frequency = 12)
approx_sl_mon_el_ts = ts(approx_sl_mon_el, start=c(1993,3), frequency = 12)
approx_at_mon_el_ts = ts(approx_at_mon_el, start=c(1993,3), frequency = 12)
approx_wt_mon_el_ts = ts(approx_wt_mon_el, start=c(1993,3), frequency = 12)


```

#### Raw Plots
```{r}
plot_ts <- function(tsdat, title, ylab, xlab) {
  autoplot(tsdat) +
    ggtitle(title) +
    ylab(ylab) +
    xlab(xlab)
    
}
plot_ts(raw_sl_mon, "Raw Monthly Sea Level Data", "Meter", "Year")
plot_ts(raw_at_mon, "Raw Monthly Air Temperature Data", "Celsius (°C)", "Year")
plot_ts(raw_wt_mon, "Raw Monthly Water Temperature Data", "Celsius (°C)", "Year")
plot_ts(raw_sl_mon_el, "Raw Monthly Sea Level Data Except Strong El-Niño", "Meter", "Year")
plot_ts(raw_at_mon_el, "Raw Monthly Air Temperature Data Except Strong El-Niño", "Celsius (°C)", "Year")
plot_ts(raw_wt_mon_el, "Raw Monthly Water Temperature Data Except Strong El-Niño", "Celsius (°C)", "Year")
```

#### Filled data plots
```{r}
plot_ts(approx_sl_mon, "Preprocssing Monthly Sea Level Data", "Meter", "Year")
plot_ts(approx_at_mon, "Preprocssing Monthly Air Temperature Data", "Celsius (°C)", "Year")
plot_ts(approx_wt_mon, "Preprocssing Monthly Water Temperature Data", "Celsius (°C)", "Year")
plot_ts(approx_sl_mon_el, "Preprocssing Monthly Sea Level Data Except Strong El-Niño", "Meter", "Year")
plot_ts(approx_at_mon_el, "Preprocssing Monthly Air Temperature Data Except Strong El-Niño", "Celsius (°C)", "Year")
plot_ts(approx_wt_mon_el, "Preprocssing Monthly Water Temperature Data Except Strong El-Niño", "Celsius (°C)", "Year")
```


### - Check if a time series is stationary: confirm the stationarity of the dataset

-   We use Dickey-Fuller test using `adf.test` function of `tseries` package. Augmented Dickey-Fuller (ADF) t-statistic test to find if the series has a unit root (a series with a trend line will have a unit root and result in a large p-value).

```{r}
options(warn=-1)
library(tseries)
adf.test(approx_sl_mon_ts)

adf.test(approx_at_mon_ts)

adf.test(approx_wt_mon_ts)

adf.test(approx_sl_mon_el_ts)

adf.test(approx_at_mon_el_ts)

adf.test(approx_wt_mon_el_ts)
```

### Check ACF
-   Autocorrelation Function (ACF) Identify if correlation at different time lags goes to 0. Just as correlation measures the extent of a linear relationship between two variables, autocorrelation measures the linear relationship between lagged values of a time series.

```{r}
library(cowplot)
library(patchwork)
iris1 <- autoplot(approx_sl_mon_ts)+
  ggtitle("Monthly Sea Level Data") +
  ylab("meter") +
  xlab("Year")

## Seasonality Plot
#ggseasonplot(approx_sl_mon_ts, year.labels=TRUE, year.labels.left=TRUE) +
#  ylab("meter") +
#  ggtitle("Seasonal plot: Monthly Sea Level")

iris2 <-ggAcf(approx_sl_mon_ts)+
  ggtitle("Autocorrelation for monthly sea level (m)")


#plot_grid(iris1, iris2, labels = "AUTO")
#iris1 + iris2

iris2

ggAcf(approx_at_mon_ts)+ 
  ggtitle("Autocorrelation for monthly air level(°C)")

ggAcf(approx_wt_mon_ts)+
  ggtitle("Autocorrelation for monthly water temperature (°C)")

ggAcf(approx_sl_mon_el_ts)+
  ggtitle("Autocorrelation for monthly sea level (m) except Strong El-Niño")

ggAcf(approx_sl_mon_el_ts)+
  ggtitle("Autocorrelation for monthly air temperature(°C) except Strong El-Niño")

ggAcf(approx_sl_mon_el_ts)+
  ggtitle("Autocorrelation for monthly water temperature(°C) except Strong El-Niño")
```
Monthly sea level is not stationary, since most lags exceed the confidence interval of the ACF (blue dashed line).

## STL model

-   t.window controls wiggliness of trend component.

-   s.window controls variation on seasonal component.
These control how rapidly the trend-cycle and seasonal components can change. Smaller values allow for more rapid changes. 

```{r stlagain, echo=TRUE, warning=FALSE, fig.width=8, fig.height=4}
sl_fit <- stl(approx_sl_mon_ts, s.window=5, robust=TRUE)
autoplot(sl_fit) +
  ggtitle("STL decomposition of monthly sea level (m)")

at_fit <- stl(approx_at_mon_ts, s.window=5, robust=TRUE)
autoplot(at_fit) +
  ggtitle("STL decomposition of monthly air temperature (°C)")

wt_fit <- stl(approx_wt_mon_ts, s.window=5, robust=TRUE)
autoplot(wt_fit) +
  ggtitle("STL decomposition of monthly water temperature (°C)")

sl_el_fit <- stl(approx_sl_mon_el_ts, s.window=13, robust=TRUE)
autoplot(sl_el_fit) +
  ggtitle("STL decomposition of monthly sea level (m) except strong El Niño.")

at_el_fit <- stl(approx_at_mon_el_ts, s.window=13, robust=TRUE)
autoplot(at_el_fit) +
  ggtitle("STL decomposition of monthly air temperature (°C) except strong El Niño.")

wt_el_fit <- stl(approx_wt_mon_el_ts, s.window=13, robust=TRUE)
autoplot(wt_el_fit) +
  ggtitle("STL decomposition of monthly water temperature (°C) except strong El Niño.")
```

```{r mstl, fig.width=8, fig.height=4}
approx_sl_mon_ts %>% mstl() %>% autoplot()

```

### Find the outliers by using STL decomposition
```{r}
plot_resid <- function(res) {
  mu = mean(res)
  sd = sd(res)
  plot(res) + 
  abline(h = c(mu-3*sd, mu+3*sd), col = c("blue","blue"), lty=c(2,2))
}
```
```{r}
## calculate the reminders in each STL decomposition
sl_el_resid <- remainder(sl_el_fit)
at_el_resid <- remainder(at_el_fit)
wt_el_resid <- remainder(wt_el_fit)

plot_resid(sl_el_resid)
plot_resid(at_el_resid)
plot_resid(wt_el_resid)
```

